Extended Data

### 2021-09-15 doc/manuscript/Dow_manuscript_Nature.Rmd
incorporate zohner_2021

Warming temperatures in recent decades have resulted in widespread phenological changes indicative of earlier arrival of spring[@parmesan_globally_2003, @ibanez_forecasting_2010,@jeong_phenology_2011,@friedl_tale_2014] and sometimes later autumn senesence[@ibanez_forecasting_2010,@jeong_phenology_2011; @keenan_net_2014].
In temperate deciduous forests, spring leaf-out has been occurring earlier[@ibanez_forecasting_2010, @jeong_phenology_2011], as has the beginning of the annual period of net CO~2~ sequestration [@piao_net_2008; @barichivich_large-scale_2013; @keenan_net_2014].
Warmer spring temperatures and earlier leaf-out are also linked to earlier leaf senescence[@zohner_ongoing_2019;@keenan_timing_2015; @fu_variation_2014], apparently because of sink-limitation to CO~2~ sequestration[@zani_increased_2020; @lempereur_growth_2015].
However, warmer summer and autumn temperatures and drought led to a delay of leaf senescence[@zohner_ongoing_2019;@xie_predicting_2018; @crabbe_extreme_2016; @dragoni_evidence_2011; @jeong_phenology_2011; @ibanez_forecasting_2010; @vitasse_responses_2009] while heavy rainfall and heat stress led to earlier senescence[@xie_predicting_2018]. As a result, warming may advance or delay autumn senescence, with delays more common across eastern North America [@buermann_widespread_2018].  
The net result is that spring warming generally results in an increased growing season length, measured as the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;;@ibanez_forecasting_2010;;@menzel_european_2006;@menzel_growing_1999] and the ecosystem is a net CO~2~ sink (Are flux sources used above ok to put here?) [@piao_net_2008; @barichivich_large-scale_2013; @keenan_net_2014].
The expectation, is that as growing season length increases, net annual carbon sequestered will increase as well (*i,e*, net ecosystem exchange, $NEE$)[@churkina_spatial_2005; @keenan_net_2014; @zhou_explaining_2016; @fu_maximum_2019], and [this expectation is represented in models](https://github.com/EcoClimLab/growth_phenology/issues/56) [@buermann_widespread_2018; @zani_increased_2020].
However, recent evidence that positive effects of warm springs are compensated by negative effects of accumulation of seasonal water deficits [@buermann_widespread_2018] and that leaf senescence and stem growth cessation is induced when water deficits reach a certain point[@lempereur_growth_2015] or sink factors bcome limiting [@zani_increased_2020] suggests that warmer spring temperatures and earlier leaf-out could advance autumn senescence and may not have the expected effect on C sequestration.

@@

As Earth's atmospheric CO~2~ concentrations and temperatures increase, the responses and feedbacks of forests will play an important yet uncertain role in influencing atmospheric CO~2~ and climate change[@arora_carbon_2020; @friedlingstein_global_2020]. 
Understanding how trees are responding to warming temperatures and rising CO2 levels, and how growth may become limited because of interactions between the two is essential to improve our predictions. 
We know very little about how the longest-lived component of fixed carbon in trees, the woody growth, is responding to warming spring temperatures. 
In fact, we know little about the stem-growth phenology of temperate deciduous species, with multi-year records of growth phenology having been published for only a few site [@doranngeville_radial_inreview; @martin_hydrometeorology_2017]. *(We need to make sure we're not missing anything. See European study cited in A-T et al in review)* 
Tree-ring records, which can be used to examine correlations of annual growth to monthly temperatures but not to understand growth phenology, reveal that growth of temperate deciduous trees tends to be most sensitive to summer temperatures (REFS), with some hints that warmer springs may have a modest positive effect on growth (e.g., Refs.[@helcoski_growing_2019;@] [MORE]).
Thus, tree-ring records do not necessarily support the suggestion that warming spring temperatures would increase annual CO~2~ sequestration. 
Characterizing phenological responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms.(Loic suggests adding numbers to illustrate the importatnce of stem growth as a carbon sink.)

@@

However, spring warming at SCBI in the study years used here did not result in increased deltaDBH or summer water stress (SPEI; Supp(?) SPEI Figure). This suggests that even when warmer springs are not associated with greater seasonal water stress, the lack of increased stem growth still prevails at this site. This could be because the woody growth at this site is already sink-limited.

@@

Our finding that interannual variation in woody growth is more strongly linked to conditions during the peak growing season than to growing season length aligns with parallel findings for NEE [@zhou_explaining_2016; @fu_maximum_2019].
However, there is also a disconnect between the woody growth responses to spring temperatures observed here and parallel studies on NEE implying that annual C sequestration increases -- at least modestly -- with spring temperatures[@keenan_net_2014] or with the length of the carbon uptake period[@zhou_explaining_2016; @fu_maximum_2019].
We show that the extra C does not go into woody growth, raising the question of where this C goes.
There are two main possibilities, which hold contrasting implications for the response of forest C balance to rising spring temperatures.
First, C that is fixed in late summer or fall may be used for woody growth that is not apparent from diameter measurements in the current year: either used for thickening of cell walls, a process that lags behind stem expansion[@cuny_woody_2015], or saved within trees as non-structural carbohydrates and used towards growth the following year[@trumbore_quantifying_2002;@richardson_seasonal_2013].
Tree-ring evidence from SCBI suggests that both processes may influence the climate sensitivity of annual woody growth, but that this dynamic is complex and unlikely to result in substantial additional growth in the following year [@helcoski_growing_2019].
*(we can easily test this: [GitHub issue 57](https://github.com/EcoClimLab/growth_phenology/issues/57))*
Second, any additional C that is not rapidly respired may be allocated, likely in the following growing season, to plant functions other than stem growth, including respiration, reproduction, foliage, fine roots, or root exudates[@trumbore_quantifying_2002]. 
Stem growth is more sensitive to water deficit than photosynthesis is[@muller_water_2011]. This seems to promote the investment of sequestered carbon into root growth in times of drought stress. Zoehner et al 2021[@zohner_how_2021], found that the proportion of root-to-stem growth increased under spring warming, suggesting that increased ecosystem sequestration resulting from spring warming may be promoting the growth of below-ground biomass.
Most of this carbon would have a relatively low residence time in the ecosystem[@trumbore_quantifying_2002], although a small fraction would be incorporated into soil organic carbon pools *with residence times rivaling that of wood [REFS]*.
Notably, both of these involve lags of a year or more and would not be captured in interannual variation in NEE, implying that the long-term effect of lengthened growing season on NEE may not be captured in studies, including this one *(This study only considered a time frame of a year or less?)*, that consider time frames of a year or less[@zhou_explaining_2016; @fu_maximum_2019].
Studies examining long-term trends in growing season length and ecosystem C uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014] -- as opposed to their interannual variation -- within or including the temperate deciduous biome showed increasing trends in both variables, suggesting that the C not allocated to current year growth may have a fairly long residence time within the ecosystem; however, the amount and fate of this carbon remains unknown.

>>>>

Warming temperatures in recent decades have resulted in widespread phenological changes indicative of earlier arrival of spring[@parmesan_globally_2003, @ibanez_forecasting_2010,@jeong_phenology_2011,@friedl_tale_2014] and sometimes later autumn senesence[@ibanez_forecasting_2010,@jeong_phenology_2011; @keenan_net_2014].
In temperate deciduous forests, spring leaf-out has been occurring earlier[@ibanez_forecasting_2010, @jeong_phenology_2011], as has the beginning of the annual period of net CO~2~ sequestration [@piao_net_2008; @barichivich_large-scale_2013; @keenan_net_2014].
Warmer spring temperatures and earlier leaf-out are also linked to earlier leaf senescence[@zohner_ongoing_2019;@keenan_timing_2015; @fu_variation_2014], apparently because of sink-limitation to CO~2~ sequestration[@zani_increased_2020; @lempereur_growth_2015].
However, warmer summer and autumn temperatures and drought led to a delay of leaf senescence[@zohner_ongoing_2019;@xie_predicting_2018; @crabbe_extreme_2016; @dragoni_evidence_2011; @jeong_phenology_2011; @ibanez_forecasting_2010; @vitasse_responses_2009] while heavy rainfall and heat stress led to earlier senescence[@xie_predicting_2018]. As a result, warming may advance or delay autumn senescence, with delays more common across eastern North America [@buermann_widespread_2018].

Spring warming generally results in an increased growing season length, measured as the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;;@ibanez_forecasting_2010;;@menzel_european_2006;@menzel_growing_1999] and the ecosystem is a net CO~2~ sink (Are flux sources used above ok to put here?) [@piao_net_2008; @barichivich_large-scale_2013; @keenan_net_2014].
The expectation, is that as growing season length increases, net annual carbon sequestered will increase as well (*i,e*, net ecosystem exchange, $NEE$)[@churkina_spatial_2005; @keenan_net_2014; @zhou_explaining_2016; @fu_maximum_2019], and [this expectation is represented in models](https://github.com/EcoClimLab/growth_phenology/issues/56) [@buermann_widespread_2018; @zani_increased_2020].

However, recent evidence that positive effects of warm springs are compensated by negative effects of accumulation of seasonal water deficits [@buermann_widespread_2018] and that leaf senescence and stem growth cessation is induced when water deficits reach a certain point[@lempereur_growth_2015] or sink factors become limiting [@zani_increased_2020; @zohner_how_2021] suggests that warmer spring temperatures and earlier leaf-out may not have a positive effect on C sequestration in trees.

@@

As Earth's atmospheric CO~2~ concentrations and temperatures increase, the responses and feedbacks of forests will play an important yet uncertain role in influencing atmospheric CO~2~ and climate change[@arora_carbon_2020; @friedlingstein_global_2020]. 
Understanding how trees are responding to warming temperatures and rising CO2 levels, and how growth may become limited because of interactions between the two is essential to improve our predictions. Zohner et al, (2020)[@zohner_how_2021] found that under controled conditions, spring warming increases biomass accumulation in trees, with a larger proportion of root compared to shoot growth. The authors also found evidence of sink limitation in autumn, even with controlled water and fertilizer application to the study trees. To our knowledge, the link between phenology and growth has never been studied in-situ.

We know very little about how the longest-lived component of fixed carbon in trees, the woody growth, is responding to warming spring temperatures. 
In fact, we know little about the stem-growth phenology of temperate deciduous species, with multi-year records of growth phenology having been published for only a few site [@doranngeville_radial_inreview; @martin_hydrometeorology_2017]. *(We need to make sure we're not missing anything. See European study cited in A-T et al in review)* 
Tree-ring records, which can be used to examine correlations of annual growth to monthly temperatures but not to understand growth phenology, reveal that growth of temperate deciduous trees tends to be most sensitive to summer temperatures (REFS), with some hints that warmer springs may have a modest positive effect on growth (e.g., Refs.[@helcoski_growing_2019;@] [MORE]).
Thus, tree-ring records do not necessarily support the suggestion that warming spring temperatures would increase annual CO~2~ sequestration. 
Characterizing phenological responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms.(Loic suggests adding numbers to illustrate the importatnce of stem growth as a carbon sink.)

@@

Spring warming at SCBI and HF in the study years used here did not result in increased $\Delta DBH$ or summer water stress (SPEI; Supp(?) SPEI Figure). This suggests that even when warmer springs are not associated with greater seasonal water stress, the lack of increased stem growth still prevails at these sites. This could be because the woody growth at these sites are already sink-limited.

@@

Our finding that interannual variation in woody growth is more strongly linked to conditions during the peak growing season than to growing season length aligns with parallel findings for NEE [@zhou_explaining_2016; @fu_maximum_2019].
However, there is also a disconnect between the woody growth responses to spring temperatures observed here and parallel studies on NEE implying that annual C sequestration increases -- at least modestly -- with spring temperatures[@keenan_net_2014] or with the length of the carbon uptake period[@zhou_explaining_2016; @fu_maximum_2019].
We show that the extra C does not go into woody growth, raising the question of where this C goes.

There are two main possibilities, which hold contrasting implications for the response of forest C balance to rising spring temperatures.
First, C that is fixed in late summer or fall may be used for woody growth that is not apparent from diameter measurements in the current year: either used for thickening of cell walls, a process that lags behind stem expansion[@cuny_woody_2015], or saved within trees as non-structural carbohydrates and used towards growth the following year[@trumbore_quantifying_2002;@richardson_seasonal_2013].
Tree-ring evidence from SCBI suggests that both processes may influence the climate sensitivity of annual woody growth, but that this dynamic is complex and unlikely to result in substantial additional growth in the following year [@helcoski_growing_2019].
*(we can easily test this: [GitHub issue 57](https://github.com/EcoClimLab/growth_phenology/issues/57))*
Second, any additional C that is not rapidly respired may be allocated, likely in the following growing season, to plant functions other than stem growth, including respiration, reproduction, foliage, fine roots, or root exudates[@trumbore_quantifying_2002]. 
Stem growth is more sensitive to water deficit than photosynthesis is[@muller_water_2011]. This seems to promote the investment of sequestered carbon into root growth in times of drought stress. Zoehner et al 2021[@zohner_how_2021], found that the proportion of root-to-stem growth increased under spring warming, suggesting that increased ecosystem sequestration resulting from spring warming may be promoting the growth of below-ground biomass. However, this study was done under tightly controlled conditions within potted plants and does not represent the environmental conditions of forests like our study does.
Most of this carbon would have a relatively low residence time in the ecosystem[@trumbore_quantifying_2002], although a small fraction would be incorporated into soil organic carbon pools *with residence times rivaling that of wood [REFS]*.
Notably, both of these involve lags of a year or more and would not be captured in interannual variation in NEE, implying that the long-term effect of lengthened growing season on NEE may not be captured in studies, including this one *(This study only considered a time frame of a year or less?)*, that consider time frames of a year or less[@zhou_explaining_2016; @fu_maximum_2019].
Studies examining long-term trends in growing season length and ecosystem C uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014] -- as opposed to their interannual variation -- within or including the temperate deciduous biome showed increasing trends in both variables, suggesting that the C not allocated to current year growth may have a fairly long residence time within the ecosystem; however, the amount and fate of this carbon remains unknown.
